Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2020-12-10 jss/EpiCompare-paper/EpiCompare-paper.Rmd
first attempt at paper introduction

---
documentclass: jss
author:
  - name: Shannon K. Gallagher
    affiliation: |
      | Biostatistics Research Branch
      | National Institute of Allergy 
      | and Infectious Diseases
    # use this syntax to add text on several lines
    address: |
      | 5603 Fishers Lane
      | Rockville, MD 20852
    email: \email{shannon.gallagher@nih.gov}
    url: http://skgallagher.github.io
  - name: Benjamin Leroy
    affiliation: |
      | Dept. of Statistics
      | Carnegie Mellon University
    address: |
      | 5000 Forbes Ave.
      | Pittsburgh, PA 15213
    email: \email{bpleroy@andrew.cmu.edu}
    url:  https://benjaminleroy.github.io/

title:
  formatted: "Time invariant analysis of epidemics with \\pkg{EpiCompare}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Time invariant analysis of epidemics with \\pkg{EpiCompare}"
  # For running headers, if needed
  short:     "\\pkg{EpiCompare}"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
classoption: shortnames
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: EpiCompare.bib  #Also not included in template by default. 
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
---

@@

\section{Introduction}\label{sec:intro}

@@

## Code formatting

@@

# \proglang{R} code {short-title="R code" #r-code}

>>>>

---
documentclass: jss
author:
  - name: Shannon K. Gallagher
    affiliation: |
      | Biostatistics Research Branch
      | National Institute of Allergy 
      | and Infectious Diseases
    # use this syntax to add text on several lines
    address: |
      | 5603 Fishers Lane
      | Rockville, MD 20852
    email: \email{shannon.gallagher@nih.gov}
    url: http://skgallagher.github.io
  - name: Benjamin LeRoy
    affiliation: |
      | Dept. of Statistics & Data Science
      | Carnegie Mellon University
    address: |
      | 5000 Forbes Ave.
      | Pittsburgh, PA 15213
    email: \email{bpleroy@andrew.cmu.edu}
    url:  https://benjaminleroy.github.io/

title:
  formatted: "Time invariant analysis of epidemics with \\pkg{EpiCompare}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Time invariant analysis of epidemics with EpiCompare"
  # For running headers, if needed
  short:     "\\pkg{EpiCompare}"
abstract: >
  We present \pkg{EpiCompare}, an \proglang{R} package that suppliments and enhance current infectious disease modeling analysis pipelines as well as to encourage comparisons across these pipelines. A major contribution of this work is the set of novel \textit{time-invariate} tools for model and epidemic comparisons - including time-invariate prediction bands. \pkg{EpiCompare} encorporates \proglang{R}'s \textit{tidy} coding style to aid it rapid and easy use. This paper provides an overview of both the tools in and intuition behind \pkg{EpiCompare} and a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
classoption: shortnames
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: EpiCompare.bib  #Also not included in template by default. 
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
---

@@

# Introduction {short-title="Intro" #sec:intro}

The recent (and currently on-going) COVID-19 global epidemic has galvanized public interest in understanding more about infectious disease modeling highlighted the usefulness of research in the area of infectious disease epidemiology. Infectious disease models typically attempt to do one or more of the following: 1) predict the spread of current and future epidemics \citep[e.g. flue prediction][]{Biggerstaff2016}, 2) analyze past and current epidemics to increase scientific knowledge \citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or project epidemic scenarios under pre-specified parameters \citep[e.g. ...][]{}. The COVID-19 epidemic highlights how all three goals are important both separately and taken as a whole.  Infectious diseases inflict enormous burdens on the world: millions of lives lost and trillions of dollars spent yearly.  Correctly analyzing and addressing these issues aids in prevention and mitigation of future outbreaks.

The current epidemic of COVID-19 also highlights that these models are just one piece of the overall analysis pipeline. University based resources like John Hopkin's and government numerical dashboards (across all levels of government) during the COVID-19 epidemic remind us that descriptive statistics and visualization can be a important first step in the process (multiple \cite{}?).} Still, rightly so, a large amount of theoretical work goes into modeling epidemics, with different models focusing at the individual / agent level, network structure or just aggregate flows (review paper \cite{}?). All placing individuals / proportions of the populations into different states (e.g. suspectible, exposed, infected, recovered, etc.). With all these models, review and comparison papers in the literature and through MIDAS (Models of Infectious Disease Agent Study) Control Center helps the individual practitioner decide the correct approach.

At the same time, analysis packages often only address a portion of the analysis pipeline. Modeling tools often don't provide easy ways to compare and assess their models on new data. Moreover, exploring and modeling epidemics require transforming and \textit{tidying} data in different ways. To fill these gaps, we present our our \proglang{R} package \pkg{EpiCompare}. Our package's primary focus is to aid and advance research in the area of comparison and assessment of epidemic \& epidemiological models. In Figure \ref{fig:pipeline}, we illustrate the data analysis pipeline of infectious diseases  as 1) data pre-processing, 2) exploratory data analysis (EDA), 3) modeling and simulating, 4) post-processing, and 5) comparison and assessment; where each previous part of the pipeline influences the next. \pkg{EpiCompare} provides tools to aids practitioners in all areas of this pipeline.

\begin{figure}[!ht]
    \centering
    \includegraphics[width = 1\textwidth]{images/pipeline1.png}
    \caption{An idealized epidemiological data analysis pipeline.}
    \label{fig:pipeline}
\end{figure}

This package also emphasizes the value of approach epidemics in a \textit{time-invariant} way, not constrained to things like a specific time scale, or initial date of the first infection. Although epidemics are by definition a process that evolves over time, epidemics, including the recent COVID-19, often need to be compared in a time-invariant way to understand the processes at play. Additionally, many tools to examine the quantity of the population in each state along the infection process (for example: quantity of suspectible vs infected vs recovered individuals) don't always as intelligently capture the natural connections between the proportion of individuals in these states. Tools in \pkg{EpiCompare} attempt to give the user the ability to extend their toolkit to evaluate epidemics to also include time-invariant approaches.

This paper is broken up into the following sections, section \ref{sec:time-invariant} motivates and showcases tools of time-invariant analysis,  section \ref{sec:overview} presents an outline of how \pkg{EpiCompare} aids a practitioner in every step of the pipeline and section \ref{sec:tour} provides a thorough demonstrating of the tools through a detailed example of a full data analysis pipeline.

SHANNON: I THINK JUST INCLUDING LOTS OF LITERATURE IN INTRO + ETC. INSTEAD OF HAVING A LITERATURE REVIEW IS BETTER.

# Motivation and tools for time-invariant analysis {short-title="Time-invariant" #sec:time-invariant}

SHANNON: I FEEL LIKE WE CAN JUST CLEAN UP TO ORIGINAL SECTION AND PUT IT BACK IN HERE

# Package tool overview {short-title="Package overview" #sec:overview}

## RMD Code formatting info

@@

@@

## RMD \proglang{R} code {short-title="R code" #r-code}

### 2020-12-10 jss/EpiCompare-paper/EpiCompare-paper.tex
first attempt at paper introduction

% To remove when JSS.cls is fixed
% https://github.com/rstudio/rticles/issues/329
\RequirePackage[2020/02/02]{latexrelease}
\documentclass[
  shortnames]{jss}

@@

\author{
Shannon K. Gallagher\\Biostatistics Research Branch\\
National Institute of Allergy\\
and Infectious Diseases \And Benjamin Leroy\\Dept. of Statistics\\
Carnegie Mellon University
}
\title{Time invariant analysis of epidemics with \pkg{EpiCompare}}

\Plainauthor{Shannon K. Gallagher, Benjamin Leroy}
\Plaintitle{Time invariant analysis of epidemics with \pkg{EpiCompare}}
\Shorttitle{\pkg{EpiCompare}}

\Abstract{
The abstract of the article.
}

@@

\Address{
    Shannon K. Gallagher\\
    Biostatistics Research Branch\\
National Institute of Allergy\\
and Infectious Diseases\\
    5603 Fishers Lane\\
Rockville, MD 20852\\
  E-mail: \email{shannon.gallagher@nih.gov}\\
  URL: \url{http://skgallagher.github.io}\\~\\
      Benjamin Leroy\\
    Dept. of Statistics\\
Carnegie Mellon University\\
    5000 Forbes Ave.\\
Pittsburgh, PA 15213\\
  E-mail: \email{bpleroy@andrew.cmu.edu}\\
  URL: \url{https://benjaminleroy.github.io/}\\~\\
  }

% Pandoc citation processing

@@

\section{Introduction}\label{sec:intro}

@@

\hypertarget{code-formatting}{%
\subsection{Code formatting}\label{code-formatting}}

@@

\section[R code]{\proglang{R} code}\label{r-code}

@@

\begin{CodeChunk}
\begin{CodeOutput}
  S                          I                     R          
S "X0 * (1 - X1 * par1 / N)" "X0 * X1  * par1 / N" "0"        
I "0"                        "X1 * (1 - par2)"     "par2 * X1"
R "0"                        "0"                   "X2"       
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fortified_net <- fortify_aggregate(epimodel_sir, states = c("s.num", "i.num", "r.num")) %>%
+   mutate(Type = "EpiModel SIR",
+          sim = as.numeric(gsub("sim", "", sim)))
\end{CodeInput}
\end{CodeChunk}

@@

{\centering \includegraphics{Figs/unnamed-chunk-17-1}

@@

\caption{\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models.}\label{fig:unnamed-chunk-17}
\end{figure}
\end{CodeChunk}

>>>>

\documentclass[
  shortnames]{jss}

@@

\author{
Shannon K. Gallagher\\Biostatistics Research Branch\\
National Institute of Allergy\\
and Infectious Diseases \And Benjamin LeRoy\\Dept. of Statistics \& Data Science\\
Carnegie Mellon University
}
\title{Time invariant analysis of epidemics with \pkg{EpiCompare}}

\Plainauthor{Shannon K. Gallagher, Benjamin LeRoy}
\Plaintitle{Time invariant analysis of epidemics with EpiCompare}
\Shorttitle{\pkg{EpiCompare}}

\Abstract{
We present \pkg{EpiCompare}, an \proglang{R} package that suppliments
and enhance current infectious disease modeling analysis pipelines as
well as to encourage comparisons across these pipelines. A major
contribution of this work is the set of novel \textit{time-invariate}
tools for model and epidemic comparisons - including time-invariate
prediction bands. \pkg{EpiCompare} encorporates \proglang{R}'s
\textit{tidy} coding style to aid it rapid and easy use. This paper
provides an overview of both the tools in and intuition behind
\pkg{EpiCompare} and a thorough demonstrating of the tools through a
detailed example of a full data analysis pipeline.
}

@@

\Address{
    Shannon K. Gallagher\\
    Biostatistics Research Branch\\
  National Institute of Allergy\\
  and Infectious Diseases\\
    5603 Fishers Lane\\
Rockville, MD 20852\\
  E-mail: \email{shannon.gallagher@nih.gov}\\
  URL: \url{http://skgallagher.github.io}\\~\\
      Benjamin LeRoy\\
    Dept. of Statistics \& Data Science\\
  Carnegie Mellon University\\
    5000 Forbes Ave.\\
Pittsburgh, PA 15213\\
  E-mail: \email{bpleroy@andrew.cmu.edu}\\
  URL: \url{https://benjaminleroy.github.io/}\\~\\
  }

@@

\section[Intro]{Introduction}\label{sec:intro}

The recent (and currently on-going) COVID-19 global epidemic has
galvanized public interest in understanding more about infectious
disease modeling highlighted the usefulness of research in the area of
infectious disease epidemiology. Infectious disease models typically
attempt to do one or more of the following: 1) predict the spread of
current and future epidemics
\citep[e.g. flue prediction][]{Biggerstaff2016}, 2) analyze past and
current epidemics to increase scientific knowledge
\citep[e.g. historical measle outbreaks][]{Neal2004}, and 3) forecast or
project epidemic scenarios under pre-specified parameters
\citep[e.g. ...][]{}. The COVID-19 epidemic highlights how all three
goals are important both separately and taken as a whole. Infectious
diseases inflict enormous burdens on the world: millions of lives lost
and trillions of dollars spent yearly. Correctly analyzing and
addressing these issues aids in prevention and mitigation of future
outbreaks.

The current epidemic of COVID-19 also highlights that these models are
just one piece of the overall analysis pipeline. University based
resources like John Hopkin's and government numerical dashboards (across
all levels of government) during the COVID-19 epidemic remind us that
descriptive statistics and visualization can be a important first step
in the process (multiple \cite{}?).\} Still, rightly so, a large amount
of theoretical work goes into modeling epidemics, with different models
focusing at the individual / agent level, network structure or just
aggregate flows (review paper \cite{}?). All placing individuals /
proportions of the populations into different states (e.g.~suspectible,
exposed, infected, recovered, etc.). With all these models, review and
comparison papers in the literature and through MIDAS (Models of
Infectious Disease Agent Study) Control Center helps the individual
practitioner decide the correct approach.

At the same time, analysis packages often only address a portion of the
analysis pipeline. Modeling tools often don't provide easy ways to
compare and assess their models on new data. Moreover, exploring and
modeling epidemics require transforming and \textit{tidying} data in
different ways. To fill these gaps, we present our our \proglang{R}
package \pkg{EpiCompare}. Our package's primary focus is to aid and
advance research in the area of comparison and assessment of epidemic \&
epidemiological models. In Figure \ref{fig:pipeline}, we illustrate the
data analysis pipeline of infectious diseases as 1) data pre-processing,
2) exploratory data analysis (EDA), 3) modeling and simulating, 4)
post-processing, and 5) comparison and assessment; where each previous
part of the pipeline influences the next. \pkg{EpiCompare} provides
tools to aids practitioners in all areas of this pipeline.

\begin{figure}[!ht]
    \centering
    \includegraphics[width = 1\textwidth]{images/pipeline1.png}
    \caption{An idealized epidemiological data analysis pipeline.}
    \label{fig:pipeline}
\end{figure}

This package also emphasizes the value of approach epidemics in a
\textit{time-invariant} way, not constrained to things like a specific
time scale, or initial date of the first infection. Although epidemics
are by definition a process that evolves over time, epidemics, including
the recent COVID-19, often need to be compared in a time-invariant way
to understand the processes at play. Additionally, many tools to examine
the quantity of the population in each state along the infection process
(for example: quantity of suspectible vs infected vs recovered
individuals) don't always as intelligently capture the natural
connections between the proportion of individuals in these states. Tools
in \pkg{EpiCompare} attempt to give the user the ability to extend their
toolkit to evaluate epidemics to also include time-invariant approaches.

This paper is broken up into the following sections, section
\ref{sec:time-invariant} motivates and showcases tools of time-invariant
analysis, section \ref{sec:overview} presents an outline of how
\pkg{EpiCompare} aids a practitioner in every step of the pipeline and
section \ref{sec:tour} provides a thorough demonstrating of the tools
through a detailed example of a full data analysis pipeline.

SHANNON: I THINK JUST INCLUDING LOTS OF LITERATURE IN INTRO + ETC.
INSTEAD OF HAVING A LITERATURE REVIEW IS BETTER.

\section[Time-invariant]{Motivation and tools for time-invariant
analysis}\label{sec:time-invariant}

SHANNON: I FEEL LIKE WE CAN JUST CLEAN UP TO ORIGINAL SECTION AND PUT IT
BACK IN HERE

\section[Package overview]{Package tool overview}\label{sec:overview}

\hypertarget{rmd-code-formatting-info}{%
\subsection{RMD Code formatting info}\label{rmd-code-formatting-info}}

@@

@@

\subsection[R code]{RMD \proglang{R} code}\label{r-code}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> trans_mat <- matrix(c("X0 * (1 - X1 * par1 / N)", "X0 * X1  * par1 / N", "0",
+                   "0", "X1 * (1 - par2)", "par2 * X1",
+                   "0", "0", "X2"), byrow = TRUE, nrow = 3)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fortified_net <- fortify_aggregate(epimodel_sir, 
+                                    states = c("s.num", "i.num", "r.num")) %>%
+   mutate(Type = "EpiModel SIR",
+          sim = as.numeric(gsub("sim", "", sim)))
\end{CodeInput}
\end{CodeChunk}

@@

{\centering \includegraphics{Figs/unnamed-chunk-16-1}

@@

\caption{\label{fig:hag-simple-sir}  Original Hagelloch SIR data (black) along with 90\% prediction band and actual simulation paths from the Simple SIR and the EpiModel SIR models.}\label{fig:unnamed-chunk-16}
\end{figure}
\end{CodeChunk}
